Semi-supervised low-rank logistic regression for high-dimensional neuroimaging datasets

### 2015-05-15 paper_nips/paper.tex
first intro finished

%\nipsfinalcopy % Uncomment for camera-ready version
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
Imaging neuroscience links human behavior to aspects of brain
biology in ever-increasing datasets.
%
Existing neuroimaging methods typically perform either discovery of
neurobiological structure or evaluate targeted hypotheses on mental tasks.
%
Modelling mental operations however hinges
on the pertinence of the assumed neurobiological structure.
%
We therefore propose to the solve the unsupervised dimensionality reduction
and supervised task classification in
an identical statistical learning problem.
%
We show that this approach yields more accurate and more interpretable
neural models of psychological tasks in a reference neuroimaging dataset.
%

\textbf{keywords}: dimensionality reduction; semi-supervised learning
bioinformatics; fMRI; systems neuroscience

@@

\section{Introduction}
%
Methods used in neuroimaging research can be grouped by discovering
neurobiological structure or revealing the neural correlates associated
with mental tasks.
To discover coherent distributions of activation structure across time,
independent component analysis (ICA; \cite{beckmann2005}) is often used
to decompose the BOLD (blood-oxygen level-dependent) signals into the
important modes of variation.
The ensuing spatial activation patterns are believed to represent
brain networks of
functionally interaction brain regions.
Similarly, sparse PCA (SPCA; \cite{varoqu2011}) has been used to
separate brain activity signals into parsimonious network components.
Thus extracted brain networks have been shown to be
manifestations of electrophysiological oscillation structure \cite{hipp15}.
Their fundamental role is
attested by continued covariation during sleep and anesthesia.
%
Network discovery is typically performed by applying ICA or SPCA on
unlabeled "resting-state" data. These capture brain dynamics
during ongoing random thought without controlled environmental input.
The rationale is that the biggest fraction of BOLD signals do
not correlate with a particular behavior, stimulus, or experimental task.

@@

On the other hand, to investigate
the neural correlates underlying mental tasks,
the general linear model (GLM; \cite{friston94}) is the dominant approach.
The contribution of
individual brain voxels is estimated
according to a design matrix of experimental tasks.
Alternatively, psychophysiological interactions (PPI; \cite{friston97}),
elucidate the functional interactions between voxels as a function
of experimental tasks.
Dynamic causal modeling (DCM; \cite{stephan04}), in turn, quantifies directed,
task-driven influences between regions
by treating the brain as a nonlinear dynamic system with unknown
neuronal states. As a last example, always more neuroimaging studies model
experimental tasks by training classification algorithms on brain signals
\cite{poldrack09decoding}.
All these methods are applied to labeled task data that capture brain dynamics
during stimulus-guided behavior.
Two important conclusions can be drawn.
First, the mentioned supervised neuroimaging analyses operate
without exception in a single-voxel space. This ignores the fact that the BOLD
signal contains coherent spatial activation patterns.
Second, existing neuroimaging analyses do not acknowledge the fact that the
task-induced changes of the BOLD signal amount to less than 5\%
of baseline activity \cite{fox07}. They do thus not exploit the high similarity
of BOLD dynamics of the human brain during experimental tasks and at rest.
Indeed, similar brain networks were observed when applying ICA on
large quantities of
rest and task data \cite{smith2009}.

@@

Both aspects can be combined into a 
semi-supervised (i.e., use task and rest data)
low-rank (i.e., perform network decomposition)
approach for neuroimaging analyses.

Yet, special interest lies in the data-driven
discovery of neurobiological structure that explains differences
in cognitive conditions. Integrating unsupervised discovery and
supervised classification should preferentially identify
neurobiological structure that allows for the best predictive models. We
propose such an approach based on conjoint gradient-descent to optimize
linear decomposition and multi-task disambiguation.

Find the most discriminative linear combination of the main direction of
variation in the data

modes of variation in the brain and the neural correlates of mental
operations have mostly been studied separately so far
their correspondence remains largely unclear

we provide principled computational framework to link these previously
unconnected domains of study.

learn a manifold / low-dimensional coordinate system
motivates

PCA, SPCA and ICA are produce compact representations of uncorrelated or
independent components

combine exploratory and confirmatory analysis steps in a single framework
allows for  overlapping response pattersn

adjust the weight assigned to each spatial pattern to maximize
discriminatory power

useful intermediate representation

integrate recently gained domain knowledge into statistical learning
problems.

Traditionally, dimensionality reduction depended on linear methods such as PCA
Autoencoder \cite{hinton06}:
one-layered learning model that computes distributed
representation from the inputs by improving reconstructions from them.
dimensionality reduction by learning the underlying manifolds
for efficient representation of the data
bottleneck
-> use AE als feature extractions
guide the estimator where to put probability mass in the inputs
AE is a common feature learning algorithm
if hidden layer linear and squared error loss, it behaves like a PCA \cite{baldi1989neural}
convex with unique global minimum
optimization objective is the same as for PCA

Most nonsupervised learning algorithms can be implemented as variants in the 
autoencoder framework. This includes decmoposition and clustering task

Linear autoencoders with tied weights and specific regularization terms
yield efficient algorithms for ICA / nonlinearity -> then closer to feed-forward neural network
\cite{le2011ica}

ICA, Sparse autoencoders, and sparse coding are mathematically equivalent
under mild conditions \cite{le2011ica}

AE can caputer global and local distribution

and SPCA (??).
-> an autoencoder can capture various different compression procedures

- reformulate the AE prbleme as a regularization boejctive

-> backpropagation: solve the unsupervised ata repsentation and supervised
problem in an identical process.
coompute the error derivates of the weights
-> gradually modified according to possible compression algorithms

without a-priori knowledge of the structure
affine transformation to compression information in the input
class of problems
architecture
statistical models

directions of greatest variance; represents each data point by its coordinates along each of these directions

LR:
these weights are then fine-tuned by the LR

Even if linear should scan a wide space of intersring model, since
theoretcially, a two-layered neural networks can learn almost any function.

learn to the optimal projection of the input data
into the main direction of variation and
then classify along these principal directions.
Vincent2010: it has been shown experimen- tally that
beginning by optimizing an unsupervised criterion, oblivious of the specific
classification problem, can actually greatly help in eventually achieving
superior performance for that classifica- tion problem.

among the models it will favor those that feature most neurobiological
pertinence

From the perspective of dictionary learning, the first layer can be
viewed as a set of basis functions whose linear combinations are learned
in the second layer \cite{olshausen96}.

project the data-generating process on the axes of main variation
autoencoder allows to reverse-engineer proerties of the data-generating
neural process \cite{olshausen96}.

learn complex properties of the brain's task response patterns

convert high-dimensional data to low-dimensional network codes

data should be more easily separable in the network spacew
low-dimensional, linear structure

equates with a shallow neural network without non-linearities

@@

>>>>

\nipsfinalcopy % Uncomment for camera-ready version
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
Imaging neuroscience links human behavior to aspects of brain
biology in ever-increasing datasets.
%
Existing neuroimaging methods typically perform either discovery of
neurobiological structure or evaluation of explicit hypotheses on mental tasks.
%
Modelling mental tasks however hinges
on the pertinence of the assumed neurobiological structure.
%
We therefore propose to solve the unsupervised dimensionality reduction
and supervised task classification in
an identical statistical learning problem.
%
We show that this approach yields more accurate and more interpretable
neural models of psychological tasks in a reference neuroimaging dataset.
%

\textbf{keywords}: dimensionality reduction; semi-supervised learning;
bioinformatics; fMRI; systems neuroscience

@@

\section{Introduction}
%
Methods used in neuroimaging research can be grouped by discovering
neurobiological structure or revealing the neural correlates associated
with mental tasks.
To discover coherent distributions of activation structure across time,
independent component analysis (ICA; \cite{beckmann2005}) is often used
to decompose the BOLD (blood-oxygen level-dependent) signals into the
important modes of variation.
The ensuing spatial activation patterns are believed to represent
brain networks of
functionally interacting brain regions.
Similarly, sparse PCA (SPCA; \cite{varoqu2011}) has been used to
separate brain activity signals into parsimonious network components.
Thus extracted brain networks have been shown to be
manifestations of electrophysiological oscillation frequencies \cite{hipp15}.
Their fundamental role is
attested by continued covariation during sleep and anesthesia.
%
Network discovery is typically performed by applying ICA or SPCA on
unlabeled "resting-state" data. These capture brain dynamics
during ongoing random thought without controlled environmental input.
The rationale is that the biggest fraction of BOLD signals do
not correlate with a particular behavior, stimulus, or experimental task.

@@

On the other hand, to investigate
the neural correlates underlying mental tasks,
the general linear model (GLM; \cite{friston94}) is the dominant approach.
The contribution of
individual brain voxels is estimated
according to a design matrix of experimental tasks.
Alternatively, psychophysiological interactions (PPI; \cite{friston97}),
elucidate the functional interactions between voxels as a function
of experimental tasks.
Dynamic causal modeling (DCM; \cite{stephan04}), in turn, quantifies directed,
task-driven influences between regions
by treating the brain as a nonlinear dynamic system with unknown
ni  neuronal states. As a last example, always more neuroimaging studies model
experimental tasks by training classification algorithms on brain signals
\cite{poldrack09decoding}.
All these methods are applied to labeled task data that capture brain dynamics
during stimulus-guided behavior.
Two important conclusions can be drawn.
First, the mentioned supervised neuroimaging analyses operate
without exception in a single-voxel space. This ignores the fact that the BOLD
signal contains coherent spatial activation patterns.
Second, existing neuroimaging analyses do not acknowledge the fact that the
task-induced changes of the BOLD signal amount to less than 5\%
of baseline activity \cite{fox07}. They do thus not exploit the high similarity
of BOLD dynamics of the human brain at rest and during experimental tasks.
Indeed, similar brain networks were observed when applying ICA on
large quantities of
rest and task data \cite{smith2009}.

@@

Both aspects can be conjointly exploited in a 
semi-supervised (i.e., use rest and task data)
low-rank (i.e., perform network decomposition)
approach to neuroimaging analysis.
%
The integration of brain-network discovery in a 
supervised classification goal should identify the
neurobiological structure that allows for the best predictive models.
%
Autoencoders suggest themselves because they can emulate
variants of most nonsupervised learning algorithms,
including PCA, SPCA, and ICA.
Autoencoders
are one-layered learning models that find local and global
representation from the inputs
by improving reconstructions from them \cite{hinton06}.
%
They behave like a PCA
in case of one linear hidden layer and a squared error loss
\cite{baldi1989neural}.
This architecture yields a convex optimization objective
with unique global minimum.
Autoencoders behave like a SPCA if shrinkage terms are added for the
matrix weights in the optimization objective.
In turn, they behave like an ICA in case of a nonlinear convex
function of the first-layer activation and tied weights \cite{le2011ica}.
These authors further demonstrated that ICA, sparse autoencoders, and 
sparse coding are mathematically equivalent
under mild conditions.
In this way, autoencoders can flexibly project the neuroimaging data
onto the axes of main variation and thus
reverse-engineer properties of the data-generating
neural process \cite{olshausen96}.

In the present investigation,
an autoencoder will be fed by (unlabeled) rest data and
integrated as bottleneck
into a low-rank logistic regression fed by (labeled) task data.
Using the chain rule in back-propagation, we can then
solve the unsupervised data representation and a supervised classification
in an identical statistical learning problem.
%
From the perspective of dictionary learning, the first layer can be
viewed as a learned set of basis functions
whose linear combinations are learned
in the second layer \cite{olshausen96}.
%
Neurobiologically, this allows 
delineating a low-dimensional manifold of brain network patterns and then 
distinguishing mental tasks
by their most discriminative linear combinations.
%
Theoretically, a big reduction of the model variance is expected by using the
resting-state autoencoder as regularization mechanisms
to put probability mass on the most neurobiological
valid members of the model space.
%
The generalization performance should consequently be improved due to 
much reduced Vapnik-Chervonenkis dimensions of the classification estimator.
%
Taken together,
the important modes of variation in brain dynamics and
the neural correlates subserving mental operations
have mostly been studied in isolation.
We provide a principled computational framework to link these previously
unconnected domains of systems neuroscience.

@@

coompute the error derivates of the weights
-> gradually modified according to possible compression algorithms
